---
title: "Final_Report"
author: "Jayden Bradbury, Alex Jacoby, Jayden Thieman, Molly Wolf"
format: pdf
editor: visual
---

```{r}
#| echo: false
#| include: false
#| label: front-matter
#| warning: false
# clean up & set default chunk options
rm(list = ls())
knitr::opts_chunk$set(echo = FALSE)

# packages
library(tidyverse) # for example
library(mosaic)    # for example
library(ggformula) # for example
library(lme4)
library(ggplot2)
library(emmeans)
library(nlme)
library(readxl)
library(tinytex) 
library(xtable)
library(knitr)
library(ggthemes)
library(GGally)
library(kableExtra)
library(emmeans)
library(Stat2Data)
library(corrplot)


# read in data 

# use this space to do any data processing you need

# load data
df <- read_xlsx("Foster_Mines_Capstone_Data_v2.xlsx")

#Pivot all cycle-related columns into long format
df_long <- df %>%
  pivot_longer(
    cols = matches("(^Cyc_[A-Z]$)|(_Cyc_[A-Z]$)"),
    names_to = "FullName",       # give it a neutral name
    values_to = "Value"
  )

#Extract the cycle (e.g., Cyc_A) and variable type (e.g., E1_MPA)
df_tidy <- df_long %>%
  mutate(
    Cycle = str_extract(FullName, "Cyc_[A-Z]$"),
    VariableType = str_remove(FullName, "_Cyc_[A-Z]$"),
    VariableType = ifelse(VariableType == "Cyc", "Cycle_time", VariableType)
  ) %>%
  select(-FullName) %>%
  pivot_wider(
    names_from = VariableType,
    values_from = Value
  ) %>%
  arrange(SampleCode, Cycle)

head(df_tidy)

#Pivot the actual cycle time columns only
df_lt <- df_tidy %>%
  pivot_longer(
    cols = matches("^Cyc_[A-Z]$"),  # Cyc_A to Cyc_N
    names_to = "Cycle_column",      # Temporary name to avoid collision
    values_to = "Num_squish"
  )

#Remove NA times
df_clean <- df_lt %>%
  filter(!is.na(Num_squish)) %>%
  select(-Cycle) %>%                #Drop the existing "Cycle" column and rename
  rename(Cycle = Cycle_column)  

df_clean <- df_clean |>
  mutate(
    ChemIndex = as.numeric(str_split_fixed(SampleCode, "_", 4)[,2]),
    Porosity = as.numeric(str_split_fixed(SampleCode, "_", 4)[,3]))

df_clean <- df_clean %>%
  group_by(SampleCode) %>%
  mutate(
    Estatic_initial = first(Estatic_MPA),  # First Estatic value per SampleCode
    Estatic_rel = Estatic_MPA / Estatic_initial  # Relative to initial
  ) %>%
  ungroup()

df_clean <- df_clean |>
  select(-E1_MPA, -tandelta, -NVP)

df_clean <- df_clean %>%
   mutate(Porosity = case_when(
     Porosity == 71 ~ "Low",
     Porosity == 81 ~ "High"
   ))
 
 df_clean <- df_clean %>%
   mutate(ChemIndex = case_when(
     ChemIndex == 121 ~ "High",
     ChemIndex == 100 ~ "Medium",
     ChemIndex == 79 ~ "Low"
   ))
 
 df_clean$Porosity <- factor(df_clean$Porosity, levels = c("Low", "High"))
 
 df_clean$ChemIndex <- factor(df_clean$ChemIndex, levels = c("Low", "Medium", "High"))
 
 df_clean <- df_clean %>%
   mutate(Cycle = case_when(
     Cycle == "Cyc_A" ~ "A",
     Cycle == "Cyc_B" ~ "B",
     Cycle == "Cyc_C" ~ "C",
     Cycle == "Cyc_D" ~ "D",
     Cycle == "Cyc_E" ~ "E",
     Cycle == "Cyc_F" ~ "F",
     Cycle == "Cyc_G" ~ "G",
     Cycle == "Cyc_H" ~ "H",
     Cycle == "Cyc_I" ~ "I",
     Cycle == "Cyc_J" ~ "J",
     Cycle == "Cyc_K" ~ "K",
     Cycle == "Cyc_L" ~ "L",
     Cycle == "Cyc_M" ~ "M",
     Cycle == "Cyc_N" ~ "N"
   ))
 
 df_clean$Cycle <- factor(df_clean$Cycle,
                          levels = c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J", "K", "L", "M", "N"))
 
 df_clean <- df_clean %>%
   filter(Num_squish <= 10000)
 df_clean <- df_clean %>%
   mutate(Time = Num_squish/freq)
 
 df_clean <- df_clean |>
   filter(Estatic_rel < 1)
 
```

## 1. Project Description

Moira Foster was tasked with designing an experiment that measured helmet foam durability over long periods of time. The experiment contained three different chemical compositions of foam with two different porosity levels, to create six types of foam. Each of these foams were compressed at different frequencies, strains, and stresses over a period of time. Dr. Foster has hired **enter group name here.** to deterimine which foam type has the most consistent static stiffness over time with impact. The sample contains 71 different foam samples across the 6 combinations.

```{=html}
<!--
*Should be in paragraph from (not bullets). Written for a 3rd party such as your client’s boss, or an external review board. As part of your description, consider the following:*

-   *Type of study \[Designed experiment, Observational study, Sample survey\] and brief description of design*
-   *Stage of the experiment \[Planning stage or Analysis stage\]*
-   *Objective of study and role of this analysis*
-   *Scope of project*
-   *Intended use for results*
-->
```

## 1.1 Research Questions

**Question 1:** Which combination of porosity and chemical index is best to predict static stiffness over time after accounting for all other variables?

## 1.2 Variables

*What is (are) possible explanatory and response variables? Relevant notes about how each is measured/recorded.*

*A table is recommended here.*

We created two new variables, a time variable that was calculated by taking the number of squishes performed on a foam sample in that current cycle and dividing it by the frequency of those squishes and a relative static stiffness variable. This was done by taking every measurement of static stiffness for a given foam sample and dividing it by the initial value of static stiffness for that foam sample. This helps with seeing how the static stiffness changes over time.

Below is a table of the variables we considered in our analysis:

```{r}
#| echo: false
#| label: tbl-variables
#| tbl-cap: "Variables used in Data Analysis"

variable.desc <- data.frame(Name = c("Sample Code", "Strain", "Relative Porosity", "Frequency", "Amplitude", "Number of Squishes", "Chemical Index", "Porosity", "Relative Static Stiffness", "Time"))
variable.desc$Type <- c("Categorical", "Numerical", "Numerical", "Numerical", "Numerical", "Numerical", "Categorical", "Categorical", "Numerical", "Numerical")
variable.desc$Notes <- c("Unique identifier for sample of foam", "Amount of strain tested on foam sample in kPa", "The measured porosity of the foam sample after the initial squish", "The frequency at which the sample foam was squished", "The amplitude at which the sample foam was squished", "Number of squishes performed on foam at given time", "Chemical index of given foam sample (Low, Medium, High)", "Porosity of given foam sample (Low, High)", "Static stiffness of given foam sample relative to the initial value of static stiffness", "Time recorded at each measurement, Number of Squishes/Frequency")
knitr::kable(variable.desc, format = "latex", booktabs = TRUE,
             col.names = c("Name", "Type", "Notes"),
             longtable = FALSE) %>%
  kable_styling( full_width = FALSE) %>%
  column_spec(3, width = "8cm") %>% 
  add_header_above(c(" " = 3))
```

```{=html}
<!--
*If design is complex, here is a good place to write a detailed description.*
-->
```

# 2. Exploratory Data Analysis (EDA)

```{=html}
<!--

*Description of important variables.* - *summary statistics* - *missing Data* - *unusual or concerning observations*

*Relevant summary figures that provide intuition for the research questions and/or inform important modeling decisions (correlation structure, etc).*

*Describe important outcomes of the EDA that the reader should notice.*
-->
```

To explore what needs to be done for modeling, we perform some exploratory data analysis.

```{r}
#| echo: false
#| label: fig-Figure1
#| warning: false 
#| fig-cap: "log(Time) vs. Relative Static Stiffness for each Foam Type"
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 3

ggplot(df_clean, aes(x = log(Time), y = Estatic_rel, group = SampleCode)) +
  geom_smooth(alpha = 0.3) +
 # stat_summary(aes(group = 1), fun = mean, geom = "line", color = "black", size = 1) +
  facet_grid(Porosity ~ ChemIndex) +
  labs(title = "Relative Static Stiffness Over Time by Foam Type",
       x = "log(Time)",
       y = "Relative Estatic") +
  theme_minimal()
# grey lines are individual measurements
```

@fig-Figure1 shows relative static stiffness over log(Time). We are using log(Time) because the relationship between time and static stiffness was not linear. See @fig-estatRel-time in the appendix for more information. From this figure, we can see that there is a negative relationship between log(Time) and relative static stiffness. It also appears that, depending on the foam type, there might be a different relationship. Some slopes are steeper than others, and some of the starting values of relative static stiffness are different than others which led us to believe that in the model we would need unique intercepts and slopes based on the foam type.

```{r}
#| echo: false
#| label: fig-chem-poro
#| warning: false 
#| fig-cap: "Chemical Index vs. Relative Static Stiffness for each level of Porosity"
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 3


ggplot(df_clean, aes(x = ChemIndex, y = Estatic_rel, fill = ChemIndex)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5) +
  facet_wrap(~ Porosity) +
  labs(title = "Estatic_rel by ChemIndex and Porosity",
       y = "Relative Static Stiffness") +
  theme_minimal()
```

In @fig-chem-poro, the relationship between chemical index and relative static stiffness is different for the two levels of foam porosity. This indicates that there is likely a significant interaction between chemical index and porosity. 
<!-- Is this lvl 2 or 1 -->

Next, we look at a correlation plot between all numerical variables.

```{r}
#| echo: false
#| label: fig-correlation
#| warning: false 
#| fig-cap: "Correlation Matrix"
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4


df_numeric <- df_clean %>%
  select(Estatic_rel, strain, Stress_kPa, Relporo, freq, amp, Num_squish, Time)  # Adjust based on your actual numeric variables

# Compute the correlation matrix
corr_matrix <- cor(df_numeric, use = "complete.obs")

# Plot the correlation matrix
corrplot(corr_matrix, method = "circle", type = "lower", tl.col = "black")
```

@fig-correlation gives us a good idea of what variables might need to be included in our final model. As we can see, number of squishes and frequency are highly correlated with relative static stiffness. This is obvious, as relative static stiffness is going to have a relationship with time as we saw in @fig-Figure1. We also see that relative porosity and strain are correlated, as well as stress and strain.

Lastly, we look at how frequency is related to relative static stiffness over time. In @fig-freq-estaticRel, we can see that for lower frequencies, there is a higher relative static stiffness level recorded. This trend is the same across the different foam types. Frequency is a possible explanatory variable that we will consider in our final model.

```{r}
#| echo: false
#| label: fig-freq-estaticRel
#| warning: false 
#| fig-cap: "Frequency vs. Relative Static Stiffness over Time"
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 3
ggplot(df_clean, aes(x = log(Time), y = Estatic_rel, color = freq)) +
  geom_line(aes(group = SampleCode), alpha = 0.7, linewidth = 1.5) +
  facet_grid(Porosity ~ ChemIndex) +
  labs(
    title = "Effect of Frequency on Estatic_rel Over Time",
    x = "log(Time)",
    y = "Relative Static Stiffness (Estatic_rel)",
    color = "Frequency"
  ) +
  theme_minimal()
```


# 3. Statistical Analysis

The linear mixed effects model is defined as:

\begin{align*}
Estatic\_rel_{ij} = a_i + b_j \log(Time_{ij}) + \epsilon_{ij}
\end{align*}
\begin{align*}
a_i = \alpha_0 + \alpha_1\text{Porosity}_i + \alpha_2 \text{ChemIndex}_i + \alpha_3\text{Frequency}_i + \alpha_4 \text{NumberOfSquishes}_i + u_i
\end{align*}
\begin{align*}
b_i = \beta_0 + \beta_1\text{Porosity}_i + \beta_2 \text{ChemIndex}_i + \beta_3\text{Frequency}_i + \beta_4 \text{NumberOfSquishes}_i + v_i
\end{align*}

where:
$\epsilon_{ij} \sim N(0, \sigma_{\epsilon}^2)$,
$u_i \sim N(0,\sigma_u^2)$,
and
$v_i \sim N(0, \sigma_v^2)$.

$Estatic\_rel_{ij}$ is the relative static stiffness of the foam for the jth measurement of the ith foam sample. $a_i$ represents the sample-specific baseline stiffness for each sample. This essentially allows each sample to have its own unique intercept. Here we let the intercepts vary based on foam porosity, chemical index, frequency, and the number of squishes performed at the given time.
$b_i \log(\text{Time}_{ij})$ is a term that allows the stiffness to change over time as time increases. Like $a_i$, $b_i$ allows each sample to have its own slope. The slopes also vary based on foam porosity, chemical index, frequency, and the number of squishes performed at the given time.
$\epsilon_{ij}$ is the irreducible or random error.

Using a mixed model like this allows us to account for both variations between all samples as well as the individual differences between foam samples.


```{r}
#| warning: false
#| label: anova-stuff
#| echo: false

library(nlme)
m1 <- lme(Estatic_rel ~ log(Time):Porosity*ChemIndex +  freq + Num_squish ,random=~1|SampleCode,data=df_clean)
anova(m1)
```


```{r}
#| warning: False
#| label: tbl-sum
#| echo: false
#| tbl-cap: "Summary Table of Estimates for Model"

sumtable <- as.data.frame(sum$coefficients)

knitr::kable(sumtable)
```


In our model we allowed for a random intercept so that each sample could have their own unique baseline and we allowed for the change that each sample would experience over time to vary uniquely 
<!-- This would be random slopes which aren't included -->
. Looking at our fixed effects we can see that chemical index is not significant.
<!-- Only discuss slope/interaction -->

Frequency is significant and has a negative relationship with relative static stiffness. We also saw that the more squishes a foam indured the lower the stiffnes became. Looking at the relationship between log(Time) and porosity for both levels of porosity the stiffness will decrease over time. Now the terms we were most interested in was the three way interaction terms between log(Time), Porosity, and Chemical Index. In R it assigned the baseline to be the interaction between log(Time), high porosity and low chemical index. Thus the estimates seen for the interaction terms are in relation to this baseline. We can see that all of these estimates are negative thus there static stiffness  decreases more over time in comparison to the baseline foam type. With the data being approximately normal from the central limit theorem, we determined anything with a t-value over $|2|$ to be significant. The t-values within the model all were significant and were kept within the final model. 

```{r}
#| echo: false
#| label: em-plot
#| warning: false
#| fig-cap: "Emmeans Comparison by Porosity*ChemIndex"
emmean <- emmeans(st_stiff_model,~ Porosity *  ChemIndex)
plot(emmean)
```

<!-- Of Slopes? -->

@em-plot shows a 95% confidence interval the Relative Static Stiffness based on the unique combination of porosity and chemical index, after accounting for all other variables. This allows us to see if there are significant differences between the different foam types. The foam type with Hight Porosity and Low Chemical Index (79) has the highest stiffness while also having a small confidence interval. That means that it will maintain a high static stiffness and also not fluctuate too much. 

<!-- ```{r} -->
<!-- #| warning: False -->
<!-- #| label: tbl-emmeans -->
<!-- #| echo: false -->
<!-- #| tbl-cap: "Estimated Marginal Means for Static Stiffness." -->

<!-- # Extract pairwise comparison results for both models -->
<!-- pairs <- as.data.frame(emmeans(st_stiff_model, ~ ChemIndex * Porosity)) -->

<!-- # Generate the table using knitr::kable -->
<!-- knitr::kable(pairs, col.names = c("ChemIndex", "Porosity", "emmean", "SE", "DF", "Lower Conf-Lvl", "Upper Conf-Lvl")) -->
<!-- ```  UNNEEDED AFter picture-->


# 4. Recommendations


**Question 1:** Which combination of porosity and chemical index is best at keeping static stiffness stable over time, after accounting for all other variables?

After performing our analysis we believe that the foam type that has a porosity of 81% and a chemical index of 79 will maintain the static stiffness the best over time.
 
<!-- Does this match eda -->

# 5. Resources

*List resources that your client might find useful*

# 6. Additional Considerations

## Study Design

<<<<<<< HEAD

## Possible Outliers

  There existed possible outliers within the data that were included within the model. These extreme values are seen to be caused by machine errors in measurement recording. The extreme values had no significant impact on the model to predict static stiffness over time, but should be considered for other response variables. 
=======
## 
>>>>>>> bc1cb2bda06f92582b232649520fc379d262e2ca

-   *Limitations to the recommendations*

  Since some of the samples are smaller the results may be less gernalizable. Controlled lab conditions may lead to different long-term performance than real world applications.
  
-   *Concerns you may have about the study; suggestions for similar studies in future*

  Make time consistent for each cycle and sample. Equal or greater sample sizes. Expand the types of chemical indexes and porosities involved in the experiment.
  
-   *Technical comments*
  May be better to use time-based response modeling instead of comparing cycles focus on set time periods. Multiple imputation or mixed-effects models would strengthen results from later-cycle data where values are missing non-randomly.

# Technical Appendix

```{=html}
<!--
*Detailed information and a copy of code and or software results. Additional graphs and supporting figures may also be placed in the appendix.*
-->
```

```{r}
#| echo: false
#| label: fig-estatRel-time
#| warning: false 
#| fig-cap: "Time vs. Relative Static Stiffness"
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 3


ggplot(df_clean, aes(x = Time, y = Estatic_rel)) +
  geom_line(aes(group = SampleCode), alpha = 0.4) +
  labs(
    title = "Relative static stiffness over time",
    x = "Time",
    y = "Relative Static Stiffness (Estatic_rel)",
  ) +
  theme_minimal()
```

Here we can see a clear nonlinear relationship between relative static stiffness and time. A logarithmic transformation is needed. See @fig-estatRel-logTime.

```{r}
#| echo: false
#| label: fig-estatRel-logTime
#| warning: false 
#| fig-cap: "log(Time) vs. Relative Static Stiffness"
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 3


ggplot(df_clean, aes(x = log(Time), y = Estatic_rel)) +
  geom_line(aes(group = SampleCode), alpha = 0.4) +
  labs(
    title = "Relative static stiffness over time",
    x = "log(Time)",
    y = "Relative Static Stiffness (Estatic_rel)",
  ) +
  theme_minimal()
```

```{r}
#| label: diag-plots
#| warning: false 
#| fig-pos: "H"
#| fig-cap: "Diagnostic Plots"
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 3

# model diagnostic plots
qqnorm(residuals(st_stiff_model),
       main = "Normal Q-Q Plot: Residuals")
qqline(residuals(st_stiff_model), col = "red")

qqnorm(ranef(st_stiff_model)$SampleCode[[1]],
       main = "Normal Q-Q Plot: Random Intercepts")  
qqline(ranef(st_stiff_model)$SampleCode[[1]], col = "red")

qqnorm(ranef(st_stiff_model)$SampleCode[[2]],
       main = "Normal Q-Q Plot: Random Slopes")  
qqline(ranef(st_stiff_model)$SampleCode[[2]], col = "red")

hist(residuals(st_stiff_model))

plot(st_stiff_model)
```



### R Script

```{r ref.label=c('front-matter', 'tbl-variables', 'fig-Figure1', 'fig-chem-poro', 'fig-correlation', 'fig-freq-estaticRel', 'Model-pvalue', 'em-plot', 'fig-estatRel-time','fig-estatRel-logTime', 'diag-plots', 'anova-stuff' )}
#| echo: true
#| eval: false

# Reprinted code chunks used previously for analysis
```
